An upper bound to gas storage and delivery via pressure-swing adsorption in porous materials

### 2020-05-22 gas-adsorption/gas-ads-body.tex
made some more of corys edits and fixed MOF names in ledgend.

\section{Introduction}
The transportation sector is dominantly powered by petroleum-based
fuels~\cite{davis2009transportation}. In the United Sates, it is responsible
for 36\% of energy-related carbon dioxide emissions \cite{useia} and 50,000
premature deaths per year associated with particulate matter and ozone
emissions~\cite{caiazzo2013air}. Natural gas and hydrogen (H$_2$) are
alternative transportation fuels that emit fewer/less pollutants than
petroleum-based fuels, and thus their widespread adoption could mitigate
climate change~\cite{mcglade2015geographical} and improve air quality for human
health. Further, as the finite global petroleum resources are declining
rapidly~\cite{sorrell2010global}, the development of technologies for the
widespread adoption of sustainable transportation fuels, such as hydrogen, is
critical.

@@

At ambient conditions, both methane and hydrogen gas possess a very low
volumetric energy density compared to gasoline. Consequently, under storage
space constraints in passenger vehicles, natural gas and hydrogen must be
densified for onboard storage to achieve a reasonable driving range on a
``full'' tank of fuel. Traditional densification approaches are liquefaction,
at cryogenic temperatures and atmospheric pressure, or compression, at room
temperature and high pressures. Both approaches require expensive
infrastructure at refilling stations and significant energy input; e.g.,
liquefaction of hydrogen consumes at least 30\% of its energy
content~\cite{bossel2003energy}. Moreover, high-pressure storage tanks are
heavy, thick-walled, and non-conformable, while cryogenic storage tanks are
bulky, expensive, and afflicted by boil-off losses~\cite{hasan2009minimizing}.

A promising approach to densify natural
gas~\cite{makal2012methane,mason2014evaluating} and
hydrogen~\cite{suh2011hydrogen,garcia2018benchmark} for vehicular storage is
through physical adsorption in nanoporous materials at room
temperature~\cite{schoedel2016role}. The internal surfaces of porous materials
attract gas molecules through van der Waals, electrostatic, etc. interactions
to achieve a higher adsorbed gas density than the bulk gas at the same
temperature and pressure, allowing for room temperature and lower-pressure
storage and alleviating many drawbacks of high-pressure storage.

For a vehicle employing a porous material to store natural gas or hydrogen, the
(volumetric) \emph{deliverable capacity} of the gas is the primary
thermodynamic property that determines the driving
range~\cite{mason2014evaluating}. The adsorbed gas storage tank delivers the
gaseous fuel to the engine via an (assumed) isothermal pressure
swing~\cite{sircar2002pressure}. The deliverable capacity (see
Fig.~\ref{fig:delcap}) is the density of the gas in the material at the storage
pressure $\pfull$ minus the residual gas that remains adsorbed at the lowest
pressure $\pempty$ such that sufficient flow is maintained to feed the engine.
For commercial feasibility, the US Department of Energy (DOE) has set
deliverable capacity targets for both adsorbed methane and hydrogen storage and
delivery for vehicles. For methane, the Advanced Research Projects
Agency--Energy (ARPA-E) set a deliverable capacity target of 315\ L STP CH$_4$
per L of adsorbent at 298 K using a 65\ bar to 5.8\ bar pressure
swing~\cite{simon2015materials}. For hydrogen, the DOE set a series of
progressive targets at five year intervals, with the ultimate target of 0.05 kg
H$_2$/L~\cite{h2targetsDOE} using a 100\ bar to 5\ bar pressure swing at a
minimum of -40 $^\circ$C~\cite{allendorf2018assessment}. Thus far, despite the
emergence of highly tunable materials with enormous surface areas, such as
metal-organic frameworks~\cite{furukawa2013chemistry}, no porous material has
met these deliverable capacity targets.

@@

\section{Gas storage \& delivery by isothermal, pressure-swing adsorption}
Consider a pressure vessel onboard a vehicle (i.e. fuel tank) packed with
porous material. At the filling stage, the tank is connected to a (pure)
gaseous reservoir at pressure $\pfull$ and allowed to equilibrate. At this
point, the adsorbed gas tank is considered full. While driving, gas desorbs
from the adsorbent to the engine/fuel cell, driven by a pressure differential.
The tank is considered depleted/empty when the pressure has dropped to
$\pempty$, the pressure at which the flow rate of gas from the tank to the
engine is insufficient. However, given $\pempty \neq 0$ (pulling vacuum),
residual gas will remain trapped in the adsorbent. Therefore, the driving range
of the vehicle is primarily determined by the \emph{deliverable capacity}. The
isothermal, volumetric deliverable capacity is an intrinsic property of the
nanoporous material and its interaction with the gas.

@@

G\'omez-Gualdr\'on \emph{et al.}~\cite{gomez2017impact} introduced a model that
accounted for gas-gas interactions via an intermolecular potential and
idealized the substrate in two different ways: (1) discrete adsorption sites
packed into an FCC lattice and (2) a volume endowed with a spatially uniform,
background potential energy field. According to molecular simulation of methane
adsorption, the ARPA-E deliverable capacity target of 315 L STP/L could be
reached in both of these idealized substrates. However, both models neglected
the space occupied by atoms of the porous material that are needed to endow the
adsorption sites/volume with the attractive potential energy.

A third body of work attempted to account for both gas-gas interactions
\emph{and} steric interactions of the gas with the atoms of the porous
material. Simulations of methane adsorption in hundreds of thousands of
explicit nanoporous crystal structures---both real and hypothetical---suggested
that the ARPA-E deliverable capacity target is
infeasible~\cite{simon2015materials}; the highest simulated methane deliverable
capacity was 196 cm$^3$~STP/cm$^3$. Confidence in this conclusion rests upon
(i) the accuracy of the intermolecular potentials describing the molecular
interactions and (ii) the sufficient sampling of material space, i.e.\ that the
structures considered are representative of the set of possible materials. To
further explore material space and address sensitivity to the potentials used,
scaling the Lennard-Jones potential well depths of material atoms to model
enhanced interactions~\cite{gomez2014exploring}, placing Lennard-Jones spheres
in a unit cell randomly to form ``pseudo-materials''~\cite{kaija2018high}, and
generating fictitious potential energy fields via a generative adversarial
network trained on zeolite structures~\cite{lee2019predicting} all generated
model substrates that failed to meet the ARPA-E methane deliverable
capacity target.

@@

The thermodynamic properties of a bulk, pure gas are characterized by an
equation of state. Of particular interest for gas storage and delivery is the
density of the gas $\rho_g(\mu,T)$ as a function of chemical potential $\mu$
and temperature $T$, which is shown for methane gas in
Fig.~\ref{fig:density-vs-mu-ch4} at $T=298$\ K. For comparison, we also show
the density of methane adsorbed into several porous materials.

@@

The deliverable capacity of gas in our homogeneous substrate is thus:
\begin{align}
    \rho_D(\V,T) &= \rho(\mufull,T) - \rho(\muempty,T),
    \label{eq:DofPhi}
    \\
    &= \rho_g(\mufull-\V,T) - \rho_g(\muempty-\V,T),
\end{align}
where $\mufull$ and $\muempty$ are the chemical potentials corresponding to the
pressures $\pfull$ and $\pempty$, respectively. Thus, $\rho_D$ for our
homogeneous substrate as a function of $\V$ is the difference between two
shifted versions of $\rho_g(\mu; T)$. Figure~\ref{fig:methane-298-D} shows the
two shifted bulk methane density curves and their difference. The horizontal
axis is the difference in molar Gibbs free energy at fixed density and
temperature between the gas within the substrate and the pure gas (\gst), which
is equal to $\V$ in our model. We see a potential $\V$ that maximizes the
deliverable capacity of our ideal substrate, balancing the need to maximize the
density at $\pfull$ against the need to minimize residual gas retained at
$\pempty$.

@@

Fig.~\ref{fig:methane-298-D} clearly shows that $\V_{opt}$ provides a maximum
$\rho_D$ over all \emph{spatially uniform}, static potential energy fields. To
qualitatively argue that the spatially uniform potential $\V_{opt}$ provides a
maximum deliverable capacity over \emph{all} (including non-spatially uniform)
static potential energy fields (a broader claim), consider a
non-spatially-uniform variation on $\V_{opt}$. We will show that the effect of
this variation is to reduce the deliverable capacity. At low adsorbed gas
densities, the lowest-energy positions will be preferentially occupied, while
at higher densities, the additional gas molecules will be forced into higher
energy locations. Thus, the mean attraction will be lowest at \pfull\ and
highest at \pempty. This will result in $\gst$ increasing monotonically with
pressure. As shown in the SI (see Sec.~\ref{sec:monotonic}), this results in
the deliverable capacity of a non-uniform potential being lower than the
deliverable capacity of a uniform potential with $\V=\gst$ for either $\pfull$
or $\pempty$. Thus, a spatially uniform potential will give the best
deliverable capacity amongst all static potential energy fields that give the
same $\gst$, and the \emph{optimal} uniform potential $\V_{opt}$ will yield a
greater deliverable capacity over any static non-uniform potential.

\begin{figure}
    \centering
    \includegraphics[width=0.95\columnwidth]{methane-298-n-vs-G}
    \caption{The deliverable capacity of methane as a function of the attractive Gibbs free energy $|\gst|$.
    Experimental deliverable capacities for several MOFs (data from Ref.~\cite{mason2014evaluating, furukawa2009storage}) are shown along with the experimental values for $\gst$ at the empty and full pressures shown as dots connected by a line.}
    \label{fig:methane-298-D}
\end{figure}

\section{Results}
Although our theoretical framework allows us to place an upper bound on the
deliverable capacity of many different gases in a rigid porous material, we
focus on the maximal deliverable capacity of methane and hydrogen gas in our
homogeneous substrate due to their application as transportation fuels. To
characterize the density of the gases, $\rho_g(p, T)$, we use experimental data
from NIST~\cite{nist}, which naturally includes quantum effects that
particularly affect hydrogen at low temperature~\cite{kumar2006quantum}. In the
context of storage onboard passenger vehicles, we compare our upper bound with
several prominent porous materials using experimental adsorption isotherms from
the literature; we also compare with deliverable capacity targets set by the US
DOE.

>>>>

\section{Introduction}
The transportation sector is dominantly powered by petroleum-based
fuels~\cite{davis2009transportation}. In the United Sates, it is responsible
for 36\% of energy-related carbon dioxide emissions \cite{useia} and $\sim$50,000
premature deaths per year associated with particulate matter and ozone
emissions~\cite{caiazzo2013air}. Natural gas and hydrogen (H$_2$) are
alternative transportation fuels that emit fewer/less pollutants than
petroleum-based fuels, and thus their widespread adoption could mitigate
climate change~\cite{mcglade2015geographical} and improve air quality for human
health. Further, as the finite global petroleum resources are declining
rapidly~\cite{sorrell2010global}, the development of technologies for the
widespread adoption of sustainable transportation fuels, such as hydrogen, is
critical.

@@

At ambient conditions, both methane and hydrogen gas possess a very low
volumetric energy density compared to gasoline. Consequently, under storage
space constraints in passenger vehicles, natural gas and hydrogen must be
densified for onboard storage to achieve a reasonable driving range on a
``full'' tank of fuel. Traditional densification approaches are liquefaction,
at cryogenic temperatures and atmospheric pressure, or compression, at room
temperature and high pressures. Both approaches require expensive
infrastructure at refilling stations and significant energy input; e.g., the
energy input to liquify hydrogen is $\sim$30\% of its energy
content~\cite{bossel2003energy}. Moreover, high-pressure storage tanks are
heavy, thick-walled, and non-conformable, while cryogenic storage tanks are
bulky, expensive, and afflicted by boil-off losses~\cite{hasan2009minimizing}.

A promising approach to densify natural
gas~\cite{makal2012methane,mason2014evaluating} and
hydrogen~\cite{suh2011hydrogen,garcia2018benchmark} for vehicular storage at
room temperature is through physical adsorption in nanoporous
materials~\cite{schoedel2016role}. The internal surfaces of porous materials
attract gas molecules through van der Waals, electrostatic, etc. interactions
to achieve a higher adsorbed gas density than the bulk gas at the same
temperature and pressure, allowing for room temperature and lower-pressure
storage and alleviating many drawbacks of high-pressure, low-temperature
storage.

For a vehicle employing a porous material to store natural gas or hydrogen, the
(volumetric) \emph{deliverable capacity} of the gas is the primary
thermodynamic property that determines the driving
range~\cite{mason2014evaluating}. The adsorbed gas storage tank delivers the
gaseous fuel to the engine via an (assumed) isothermal pressure
swing~\cite{sircar2002pressure}. The deliverable capacity (see
Fig.~\ref{fig:delcap}) is the density of the gas in the material at the storage
pressure $\pfull$ minus the residual gas that remains adsorbed at the lowest
pressure $\pempty$ such that sufficient flow is maintained to feed the engine.
For commercial feasibility, the US Department of Energy (DOE) has set
deliverable capacity targets for both adsorbed methane and hydrogen storage and
delivery for vehicles. For methane, the Advanced Research Projects
Agency--Energy (ARPA-E) set a deliverable capacity target of 315\ L STP CH$_4$
per L of adsorbent at 298 K using a 65\ bar to 5.8\ bar pressure
swing~\cite{simon2015materials}. For hydrogen, the DOE set a series of
progressive targets at five year intervals, with the ultimate target of 0.05 kg
H$_2$/L~\cite{h2targetsDOE} using a 100\ bar to 5\ bar pressure swing at a
minimum of -40 $^\circ$C~\cite{allendorf2018assessment}. Thus far, despite the
emergence of highly tunable materials with large surface areas, such as
metal-organic frameworks~\cite{furukawa2013chemistry}, no porous material has
met these deliverable capacity targets.

@@

\section{Gas storage \& delivery by isothermal, pressure-swing adsorption}
Consider a pressure vessel onboard a vehicle (i.e. fuel tank) packed with
porous material. At the filling stage, the tank is connected to a (pure)
gaseous reservoir at pressure $\pfull$ and allowed to equilibrate. At this
point, the adsorbed gas tank is considered full. While driving, gas desorbs
from the adsorbent to the engine/fuel cell, driven by a pressure differential.
The tank is considered depleted/empty when the pressure has dropped to
$\pempty$, the pressure at which the flow rate of gas from the tank to the
engine is insufficient. However, given $\pempty \neq 0$ (pulling vacuum),
residual gas will remain trapped in the adsorbent. Therefore, the driving range
of the vehicle is primarily determined by the \emph{deliverable capacity} of
the gas in the material (see Fig.~\ref{fig:delcap}): the density of gas at
$\pfull$ minus that at $\pempty$. The isothermal, volumetric deliverable
capacity is an intrinsic property of the nanoporous material and its
interaction with the gas.

@@

G\'omez-Gualdr\'on \emph{et al.}~\cite{gomez2017impact} introduced a model that
accounted for gas-gas interactions via an intermolecular potential and
idealized the substrate in two different ways: (1) discrete adsorption sites
packed into an FCC lattice and (2) a volume endowed with a spatially uniform,
background potential energy field. According to molecular simulation of methane
adsorption, the ARPA-E deliverable capacity target of 315 L STP/L could be
reached in both of these idealized substrates. However, both models neglected
the space occupied by atoms of the porous material that are needed to endow the
adsorption sites/volume with the attractive potential energy and rely on a
molecular model for methane.

A third body of work incorporated both gas-gas interactions \emph{and} steric
interactions of the gas with the atoms of the porous material. Simulations of
methane adsorption in hundreds of thousands of explicit nanoporous crystal
structures---both real and hypothetical---suggested that the ARPA-E deliverable
capacity target is infeasible~\cite{simon2015materials}; the highest simulated
methane deliverable capacity was 196 cm$^3$~STP/cm$^3$. Confidence in this
conclusion rests upon (i) the accuracy of the intermolecular potentials
describing the molecular interactions and (ii) the sufficient sampling of
material space, i.e.\ that the structures considered are representative of the
set of possible materials. To further explore material space and address
sensitivity to the intermolecular potentials, scaling the Lennard-Jones
potential well depths of material atoms to model enhanced
interactions~\cite{gomez2014exploring}, placing Lennard-Jones spheres in a unit
cell randomly to form ``pseudo-materials''~\cite{kaija2018high}, and generating
fictitious potential energy fields via a generative adversarial network trained
on zeolite structures~\cite{lee2019predicting} all generated model substrates
that failed to meet the ARPA-E methane deliverable capacity target.

@@

The thermodynamic properties of a bulk, pure gas are characterized by an
equation of state. Of particular interest for gas storage and delivery is the
density of the gas $\rho_g(\mu,T)$ as a function of chemical potential $\mu$
and temperature $T$, is shown for methane gas in
Fig.~\ref{fig:density-vs-mu-ch4} at $T=298$\ K. For comparison, we also show
the density of methane adsorbed into several porous materials.

@@

The deliverable capacity of gas in our homogeneous substrate is thus:
\begin{align}
    \rho_D(\V,T) &= \rho(\mufull,T) - \rho(\muempty,T),
    \label{eq:DofPhi}
    \\
    &= \rho_g(\mufull-\V,T) - \rho_g(\muempty-\V,T),
\end{align}
where $\mufull$ and $\muempty$ are the chemical potentials corresponding to the
pressures $\pfull$ and $\pempty$, respectively. Thus, $\rho_D$ for our
homogeneous substrate as a function of $\V$ is the difference between two
shifted versions of $\rho_g(\mu; T)$. Figure~\ref{fig:methane-298-D} shows the
two shifted bulk methane density curves and their difference. The horizontal
axis is the difference in molar Gibbs free energy, at fixed density and
temperature, between the gas within the substrate and the pure gas (\gst), which
is equal to $\V$ in our model. We see a potential $\V$ that maximizes the
deliverable capacity of our ideal substrate, balancing the need to maximize the
density at $\pfull$ against the need to minimize residual gas retained at
$\pempty$.

@@

Fig.~\ref{fig:methane-298-D} clearly shows that $\V_{opt}$ provides a maximum
$\rho_D$ over all \emph{spatially uniform}, static potential energy fields. To
qualitatively argue that the spatially uniform potential $\V_{opt}$ provides a
maximum deliverable capacity over \emph{all} (including non-spatially uniform)
static potential energy fields (a broader claim), consider a
non-spatially-uniform variation on $\V_{opt}$. The effect of
this variation is to reduce the deliverable capacity: At low adsorbed gas
densities, the lowest-energy positions will be preferentially occupied, while
at higher densities, the additional gas molecules will be forced into higher
energy locations. Thus, the mean attraction will be lowest at \pfull\ and
highest at \pempty. As a result, $\gst$ increases monotonically with
pressure. As shown in the SI (see Sec.~\ref{sec:monotonic}), this results in
the deliverable capacity of a non-uniform potential with a given $\gst$ at $\pfull$ or $pempty$ being lower than the
deliverable capacity of a uniform potential with $\V=\gst$. Thus, a spatially uniform potential gives the maximal
deliverable capacity amongst all static potential energy fields that give the
same $\gst$, and the \emph{optimal} uniform potential $\V_{opt}$ will yield a
greater deliverable capacity over any static, non-uniform potential.

\begin{figure}
    \centering
    \includegraphics[width=0.95\columnwidth]{methane-298-n-vs-G}
    \caption{The deliverable capacity of methane as a function of the attractive Gibbs free energy $|\gst|$.
    Experimental deliverable capacities for several porous materials (data from Ref.~\cite{mason2014evaluating, furukawa2009storage}) are shown along with the experimental values for $\gst$ at the empty and full pressures shown as dots connected by a line.}
    \label{fig:methane-298-D}
\end{figure}

\section{Results}
Although our theoretical framework allows us to place an upper bound on the
deliverable capacity of many different gases in rigid porous materials, we
focus on the maximal deliverable capacity of methane and hydrogen gas in our
homogeneous substrate due to their application as transportation fuels. To
characterize the density of the gases, $\rho_g(p, T)$, we use experimental data
from NIST~\cite{nist}, which naturally includes quantum effects that
particularly affect hydrogen at low temperature~\cite{kumar2006quantum}. In the
context of storage onboard passenger vehicles, we compare our upper bound with
several prominent porous materials using experimental adsorption isotherms from
the literature; we also compare with deliverable capacity targets set by the US
DOE.
